---
title: "Downloading data from movebank"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading data from movebank}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- units::as_units(difftime(Sys.time(), now))
      # return a character string to show the time
      paste(
        '<div style="text-align: right; margin-top: -10px;"><font size="-3">This code took:',
        format(res, digits = 3, nsmall = 1), "</font></div>"
      )
    }
  }
}))
krba <- getOption("keyring_backend")
options("keyring_backend" = "env")
```

```{r setup}
library(move2)
```

# User credentials

The credentials of the user are stored using the `keyring` package. With the
following command a user can be added to the keyring. Run this line once, it
will store your credentials in keyring. After that every time you load `move2`
and execute a download function from movebank, these functions will retrieve
your credentials from keyring.

```{r, eval=F}
movebank_store_credentials("myUserName", "myPassword")
```

```{r, echo=F}
movebank_store_credentials("asf", "adsf", force = TRUE)
```

```{r}
movebank_remove_credentials()
```

The `keyring` package can use several mechanisms to store credentials, these are
called backends. Some of these backends are operating system dependent, others
are more general. Some of the operating systems dependent backends have the
advantage that they do not require providing credentials when opening a new R
session.

The `move2` package uses the default backend as is returned by
`keyring::default_backend()`, this function thus shows the backend `move2` is
using. If you want to change the default you can use the `keyring_backend`
option, for more details see the \code{\link[keyring]{backends}} documentation
in the keyring package.

**macOS** and **Windows** generally do not require entering an extra password
for keyring. The default in **Linux** is often the `file` backend which can be
confusing as it creates an encrypted file with credentials that need a password
to unlock. In this case a separate password for the keyring file has to be
entered for each new R session before the movebank password can be accessed. To
avoid having to enter each time a keyring password the Secret Service API can be
used by installing the `libsecret` library. (Debian/Ubuntu: `libsecret-1-dev`;
Recent RedHat, Fedora and CentOS systems: `libsecret-devel`)

###  Handling multiple Movebank accounts - use `key_name`

If you have multiple user accounts on movebank, the easiest way is to give each
of them a key name with the argument `key_name`. For the most used account also
the default option can be used. The `movebank_store_credentials()` only has to
be executed once for each account. After that the credentials will be retrieved
from keyring.

```{r, eval=F}
## store credentials for the most used account.
movebank_store_credentials("myUserName", "myPassword")

## store credentials for another movebank account
movebank_store_credentials("myUserName_2", "myPassword_2", key_name = "myOtherAccount")
```

```{r credentials, eval=TRUE, echo=FALSE}
movebank_store_credentials("myUserName", "myPassword", force = TRUE)
movebank_store_credentials("myUserName_2", "myPassword_2",
  key_name = "myOtherAccount", force = TRUE
)
```

When you want to download from Movebank using your default movebank account,
nothing has to be specified before the download functions. If you want to
download from Movebank with another account, than you should execute the line
below, specifying the key name of the account to use, before the download
functions are executed.

```{r option_setting}
options("move2_movebank_key_name" = "myOtherAccount")
```

If in one script/Rsession you are using several accounts, to use the credentials
of the default account execute the line below:

```{r options_setting_2}
options("move2_movebank_key_name" = "movebank")
```

To check which accounts are stored in keyring:

```{r, eval=FALSE}
keyring::key_list()
#   service           username
# 1 movebank          myUserName
# 2 myOtherAccount    myUserName_2
```

The `service` column corresponds to the names provided in `key_name`. The
account entered without a key name (the default) will be called `movebank`. Note
that the key names have to be unique, if there are several usernames with the
same key name (service), it will cause an error.

### Removing user credentials from keyring

To deleted credentials from keyring:

```{r remove_credentials}
## for the default account
movebank_remove_credentials()

## for an account with a key name
movebank_remove_credentials(key_name = "myOtherAccount")
```

Next we can check if the keys are successfully removed:

```{r, eval=FALSE}
keyring::key_list()
```

Here you can check if the `movebank` service is successfully removed.

# Downloading data

```{r, echo=FALSE}
options("keyring_backend" = krba)
if (Sys.info()["user"] != "bart") {
  if (Sys.getenv("MBPWD") != "") {
    options(keyring_backend = "env")
    move2::movebank_store_credentials("move2_user", Sys.getenv("MBPWD"))
  } else {
    knitr::opts_chunk$set(eval = FALSE)
  }
}
```

```{r, message=FALSE}
library(dplyr)
```

Using the `movebank_retrieve` function it is possible to directly access the
API, here all studies with a creative commons 0 license are returned. These are
a good candidate for exploration and testing

```{r cc_studies, time_it=TRUE}
movebank_retrieve(entity_type = "study", license_type = "CC_0") |>
  select(id, name, number_of_deployed_locations) |>
  filter(!is.na(number_of_deployed_locations))
```

A more quick way to retrieve the information is the following (the selection is
performed on movebank and not all data is downloaded):

```{r studies, eval=FALSE}
movebank_download_study_info(license_type = "CC_0")
```

By default all attributes are downloaded:

```{r full_dl, time_it=TRUE}
movebank_download_study(2911040, sensor_type_id = "gps")
```

For speed of download you might want to add the argument `attributes = NULL` as
it reduces the columns to download to the bare minimum. Note still all
individual attributes are downloaded as this does not take much time.

```{r quick_download, time_it=TRUE}
movebank_download_study(1259686571, sensor_type_id = "gps", attributes = NULL)
```

If only specific attributes want to be download you can state them in the argument `attributes`. The available attributes vary between studies and sensors. You can retrieve the list of available attributes for a specific sensor in given study. Note that only one sensor at a time can be stated.
```{r study_attrs, time_it=T}
movebank_retrieve(
  entity_type = "study_attribute",
  study_id = 2911040,
  sensor_type_id = "gps"
)$short_name
movebank_download_study(
  study_id = 2911040,
  sensor_type_id = "gps",
  attributes = c(
    "height_above_ellipsoid",
    "eobs_temperature"
  )
)
```


Only load gps records:

```{r gps_sensor, time_it=TRUE}
movebank_download_study(1259686571, sensor_type_id = 653)
```

Note that the `sensor_type_id` can either be specified either of an `integer` or
`character` with respectively the id or name of the sensor. In some cases
additional data is added is downloaded if a specific sensor is selected. For
example the column `eobs_acceleration_raw`:

```{r acc_study_download, time_it=TRUE}
movebank_download_study(2911040, sensor_type_id = "acceleration")
```

The following list of sensors is available:

```{r retrieve_sensors}
movebank_retrieve(
  entity_type = "tag_type",
  attributes = c("external_id", "id")
)
```

Alternatively more informative names can be used for some arguments. For example
you can use a `character` string to identify a study or a timestamp as a
`POSIXct`:

```{r download_lbbg, time_it=TRUE}
movebank_download_study("LBBG_JUVENILE",
  sensor_type_id = "gps",
  timestamp_start = as.POSIXct("2021-02-03 00:00:00"),
  timestamp_end = as.POSIXct("2021-03-03 00:00:00")
)
```

# Deployments

If you are interested in the deployment information you can use the
`movebank_download_deployment` function.

```{r galapagos_deployment, time_it=T}
movebank_download_deployment("Galapagos Albatrosses")
```

# Advanced usage

For specific request it might be useful to directly retrieve information from
the movebank api. The `movebank_retrieve` function provides this functionality.
The first argument is the entity type you would like to retrieve information for
(e.g. `tag` or `event`). Other arguments make it possible to select, a study id
is always required. For more details how to use the api see the
[documentation](https://github.com/movebank/movebank-api-doc/blob/master/movebank-api.md).

## Downloading undeployed data
One common reason to use this options is to retrieve undeployed locations. In
some cases a set of locations is collected before the tag attached to the animal
for quality control or error measurements. The example below shows how all
records for a specific tag can be retrieved. Filtering for locations where the
`deployment_id` is `NA`, returns those locations that were collected while the
tag was not deployed. The `timestamp_start` and `timestamp_end` might be good
argument to filter down the data even more in the call to `movebank_retrieve`. 
By omitting the argument `tag_local_identifier` the entire study can downloaded. 
With the argument `sensor_type_id` the sensors can be specified.

```{r advance, time_it=TRUE}
movebank_retrieve("event",
  study_id = 1259686571,
  tag_local_identifier = "193967", attributes = "all"
) %>%
  filter(is.na(deployment_id))
```
